---
title: "Converting Dates to Periods, Cohorts, Age Groups, and Lexis Triangles"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Converting Dates to Periods, Cohorts, Age Groups, and Lexis Triangles}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
library(demprep)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  fig.width = 6.5,
  fig.align = "center"
)
deaths <- data.frame(name = c("Anwar", "Baptiste", "Candice"),
                     date_birth = c("2014-02-17", "2012-01-10", "2019-04-29"),
                     date_death = c("2019-10-11", "2020-02-27", "2020-08-01"))

births_in <- data.frame(date_child = c("2020-05-18", "2020-07-02", "2020-10-05"),
                        date_parent = c("2000-02-14", "2002-05-18", "1990-12-18"))

births_out <- data.frame(date_child = c("2020-03-14", "2020-08-22"),
                         date_parent = c("1968-01-23", "2008-03-05"))
```

# Introduction

Demographic datasets typically classify events using age groups and periods, like the table below:
```{r deaths-array, echo = FALSE}
deaths_array <- matrix(c(98, 25, 11, 89, 29, 12, 93, 32, 16),
                       nrow = 3, 
                       ncol = 3,
                       dimnames = list(age = paste("Age", c("0", "1-4", "5-9")),
                                       year = c("2000-2005", "2005-2010", "2010-2015")))
kable(deaths_array,
      align = "c",
      caption = "Counts of deaths by age group and year")
```
The demographic datasets ultimately derive from individual-level data contains specific dates, rather than age groups, like the data below:
```{r deaths-list, echo = FALSE}
kable(deaths,
      caption = "Individual-level data on births and deaths")
```

Converting dates into age groups and periods requires care, and can sometimes be difficult. Consider, for instance, the problem of measuring a person's age in months. If a person is born on 31 January 2001, on what date does that person become one month old? On 28 February 2001, on 1 March 2001, or on some other date? According to base *R*, the answer is 3 March 2001:
```{r one-month}
seq.Date(from = as.Date("2001-01-31"), 
         by = "month",
         length.out = 2)
```

Package **demprep** contains a suite of functions for converting dates to age groups, periods, and other age-time categories needed by the **dem** packages. These functions deal with details such as whether one-year periods start in January or July. It also follows a consistent approach to dealing with the fact that months and years have variable lengths. Under this approach, the entry into a new age group, or the start of a new period, depends on reaching a certain day of the month or month of the year, rather than on the number of days that have elapsed. As we discuss below, this approach reflects common practices in applied demography.

DOESN'T TRY TO DEAL WITH ALL POSSIBLE SCENARIOS, SCHEMES - TO AVOID OVERLY COMPLICATED INTERFACE.
- a bit prescriptive. Can always use general-purpose R tools, eg 'cut'. 

# Age-time plans

```{r agetime, echo = FALSE}
date <- as.Date(c("2020-12-07", "2022-05-30"))
dob <- as.Date(c("2001-07-22", "2006-03-14"))
age <- c(19 + 4.5/12, 16 + 2.5/12)
```

When reasoning about age and time, demographers often make use of Lexis diagrams. The figure below is an example of a Lexis diagram. The horizontal axis measures calendar time, and the vertical axis measures age. The black dots marked $A$ and $B$ represent deaths. The black diagonal lines are "life lines" for the people experiencing these deaths. The black dots on the horizontal axis represent births. A life line in a traditional Lexis diagram always covers the same distance along the vertical axis as it does along the horizontal axis, because a person always gains one year of age for each year of calendar time.

```{r lexis1, echo = FALSE, fig.width = 5, fig.height = 5}
breaks_time <- seq.Date(from = as.Date("2000-01-01"),
                        to = as.Date("2025-01-01"), 
                        by = "5 years")
breaks_age <- seq.int(from = 0L, 
                      to = 25L, 
                      by = 5L)
year <- as.integer(format(breaks_time, "%Y"))
labels_time <- demprep:::make_labels_period_custom(year,
                                                   include_na = FALSE)
labels_age <- demprep:::make_labels_age(breaks = breaks_age,
                              open_last = FALSE,
                              include_na = FALSE)
demprep:::plot_date_to_age_triangle(date = date,
                                    dob = dob,
                                    unit = "year",
                                    breaks_time = breaks_time,
                                    breaks_age = breaks_age,
                                    labels_time = labels_time,
                                    labels_age = labels_age,
                                    show_vert = TRUE,
                                    show_diag = TRUE,
                                    show_months = FALSE)
text(x = date,
     y = age,
     labels = expression(italic(A), italic(B)),
     pos = 4)
```

The light blue grid in the Lexis diagram above depicts an "age-time plan" [references]. An age-time plan is a set of rules for how exact times and ages (represented by dots in the Lexis diagram) are assigned to discrete age and time categories. The age-time plan above divides divides age into 5-year age groups, and divides time into 5-year periods, with each period starting on 1 January. Deaths *A* and *B* both belong to age group "15-19" and period "2020-2025".

Although age group and period are the most commonly-used age-time categories in demography, some applications also make use of cohorts. A cohort is a group of people who all experience a particular event during a given period. The event is normally birth, but it can be, for instance, marriage or joining an organization. The people who are born during the period 1950-1960, form a birth cohort, for example, and the people who marry during the year 2018 form a marriage cohort.

On a Lexis diagram, the birth cohort for a person or event can be inferred by following the relevant life line back to the horizontal axis and noting which period the birth falls within. In the Lexis diagram above, for instance, death *A* belongs to the birth cohort "2000-2005" and death *B* belongs to the birth cohort "2005-2010".

When working with cohorts, it is often helpful to add one additional age-time category: the Lexis triangle. Each combination of age group and period has two Lexis triangles. Death *A* in the diagram above belongs to an upper Lexis triangle, and death *B* belongs to a lower Lexis triangle.

Knowing the age group and period of an event is not enough to fully determine the cohort for that event. If we did not have access to the Lexis diagram above, for instance, and all we knew about death *A* was that it belonged to age group "15-19" and period "2020-2025", then we would not be able to determine whether *A* belonged to the the cohort "2000-2005" or to the cohort "2005-2010". If, however, we also knew that *A* belonged to an upper Lexis triangle, then the ambiguity would be be removed: we could infer that *A* belonged to the cohort "2000-2005". Similarly, if we knew that *B* belonged to the lower Lexis triangle within the square formed by age group "15-19" and period "2020-2025", we could conclude that *B* belongs to the cohort "2005-2010".

Lexis triangles are only defined when age groups and periods have the same length, as in the diagram above, when both have lengths of 5 years. Age-time plans with unequal lengths are, in fact common, such as age-time plans with 5-year age groups and 1-year periods.

**Measuring periods using using dates attained, rather than days elapsed***

In everyday usage, the endpoints of a period depend on dates attained rather than days elapsed. The period from 1 January 2020 to 1 January 2021 and the period from 1 January 2021 to 1 January 2022 both count as one year, even though the first period has 366 days and the second has 365. Similarly, the period from 1 June to 1 July and the period from 1 July to 1 August both count as one month, even though the first period has 30 days and the second has 31.

In the case of years, periods are defined by lining up the days of the month and the month of the year, and then calculating at the difference in years. For instance, 30 June 2014 and 30 June 2017 are suitably aligned, so we can conclude that the two dates are 3 years apart. 

In the case of months, periods are defined by lining up days of the month and then calculating the difference month. For instance, 1 January 2000 and 1 April 2001 are suitably aligned, so we can conclude that the two dates are 15 months (or, equivalently, one year and three months) apart.

Sometimes, however, it is not possible to align days and months. If we want to go one year forward from 29 February 2020, we there is no 29 February 2021 to align to. As noted above, a similar problem arises when trying to go one month forward from 31 January in any year. These sorts of problems arise for days of the month, such as the 31st, that occur in some months but not others.

**demprep** sidesteps the problem of irregular days of the month by requiring that all periods start on the first of the month. For example, year-long periods can begin on any month of the year, but they must always begin on the first day of the month. A period beginning on 1 December is allowed, but a period beginning on 2 January is not. This restriction makes the practice of defining periods simpler and more predictable, and is consistent with what most producers of data do in any case.





# Calculating ages, periods, cohorts, and Lexis diagrams from data on dates 

| Periods | Cohorts | Age groups | Lexis triangles |
|:--------|:--------|:-----------|:----------------|
| `date_to_period_year` | `date_to_cohort_year` | `date_to_age_year` | `date_to_triangle_year` |
| `date_to_period_quarter` | `date_to_cohort_quarter` | `date_to_age_quarter` | `date_to_triangle_quarter` |
| `date_to_period_month` | `date_to_cohort_month` | `date_to_age_month` | `date_to_triangle_month` |

The last part of the name of each function refers to units that are created.


## Periods

We begin with function `date_to_period_year`, which assigns dates to 1-year periods. We illustrate this, and other functions, using the dataset introduced above:
```{r deaths}
deaths
```
By default, `date_to_period_year` creates periods that start on 1 January. To demonstrate, using a dataset, introduced above, called `deaths`:
```{r}
deaths
```
We make a temporary copy of `deaths`, and then call `date_to_period_year` on the `date_death` column, using the default settings for `date_to_period_year`. By default, `date_to_period_year` constructs periods that start on 1 January:
```{r}
deaths_temp <- deaths
deaths_temp$period_jan <- date_to_period_year(date = deaths_temp$date_death)
deaths_temp
```
The two periods created by `date_to_period_year` are depicted below. 
```{r, echo = FALSE, fig.height = 1.75}
demprep:::plot_date_to_period_year(date = deaths$date_death)
```
The first starts on 1 January 2019, and the second starts on 1 January 2020.

To create periods that start on a date other than 1 January, we use the `month_start` argument:
```{r}
deaths_temp$period_jul <- date_to_period_year(date = deaths_temp$date_death,
                                              month_start = "Jul")
deaths_temp
```
The resulting periods start on 1 July 2019 and 1 July 2020:
```{r, echo = FALSE, fig.height = 1.75}
demprep:::plot_date_to_period_year(date = deaths_temp$date_death,
                         month_start = "Jul")
```

The first of these periods spans calendar years 2019 and 2020, and the second spans calendar years 2020 and 2021. By default, `date_to_period_year` labels single-year periods using the calendary year at the start of the period. The periods above are therefore labelled "2019" and "2020". Authorities follow a different convention, and refer to the periods above as the year to June 2020, and the year to June 2021, leading to the labels "2020" and "2021". To create labels based on the calendar year at the end of each period, rather than the calendar year at the begining, we set the `label_year_start` argument of `date_to_period_year` to `FALSE`:
```{r}
deaths_temp$period_jul_end <- date_to_period_year(date = deaths_temp$date_death,
                                                  month_start = "Jul",
                                                  label_year_start = FALSE)
deaths_temp
```

The resulting periods are defined by exactly the same dates as the previous ones, but have different labels.
```{r, echo = FALSE, fig.height = 1.75}
demprep:::plot_date_to_period_year(date = deaths_temp$date_death,
                         month_start = "Jul",
                         label_year_start = FALSE)
```

When one-year periods start on 1 January, the question of whether to label by the start year or the end year does not arise, since the period falls within a single calendar year.

`date_to_period_year` creates enough periods so that every element of the `date` argument can be assigned to a period. As the following example demonstrates, `date_to_period_year` also, by default, fills in any gaps between the first and last period, even if the resulting periods do not contain any elements from `date`. We first create a dataset with no deaths during the year 2018.
```{r}
deaths_temp <- rbind(deaths,
                     data.frame(name = "Ding", 
                                date_birth = "2017-03-18",
                                date_death = "2017-07-15"))
deaths_temp$period <- date_to_period_year(date = deaths_temp$date_death)
deaths_temp
```


Quarters are three-month periods starting on the first day of January, April, July, and October:
```{r}
deaths_temp <- deaths
deaths_temp$time <- date_to_period_quarter(deaths_temp$date_death)
deaths_temp
```

```{r, echo = FALSE, fig.height = 1.75}
demprep:::plot_date_to_period_quarter(deaths_temp$date_death)
```

Month-length periods are similarly straightforward:
```{r}
deaths_temp <- deaths
deaths_temp$time <- date_to_period_month(deaths_temp$date_death)
deaths_temp
```

```{r, echo = FALSE, fig.height = 1.75}
demprep:::plot_date_to_period_month(deaths_temp$date_death[1:2])
```


## Ages

Abstract: $\text{age} = t_{\text{event}} - t_{\text{dob}}$

Complications. 

- Child born on 31 October. On what day does child become 1 month old? 
- Child born on 30 November. On what day does child become 1 month old?
- Child born on 29 February 2000. On what day does child become 1 year old? [1 March 2001]

Principles:
- tick over when day of date equals day of dob. Includes leap years.

\begin{equation}
\begin{split}
\text{age in completed months} & = 12 \times (\text{year of event} - \text{year of birth}) \\
& \quad + \text{month of event} - \text{month of birth} \\
& \quad - \text{partial-month adjustment}
\end{split}
\end{equation}

Partial-month adjustment: if $\text{day of event} \ge \text{day of birth}$, then 0; otherwise 1.


```{r, echo = FALSE, fig.height = 6.5}
date <- as.Date(c("2021-03-30", "2021-03-10"))
dob <- as.Date(c("2021-01-30", "2021-01-10"))
demprep:::plot_date_to_age_month(date = date,
                       dob = dob)
```

\begin{equation}
\text{age in completed quarters} = \text{floor}(\text{age in completed months} / 3)
\end{equation}

\begin{equation}
\text{age in completed quarters} = \text{floor}(\text{age in completed months} / 12)
\end{equation}


| Relationship between dates | Age in completed years |
|:---------------------------|:-----------------------|
| $\text{month of event} > \text{month of birth}$ | $\text{year of event} - \text{year of birth}$ |
| $\text{month of event} = \text{month of birth}$, $\text{day of event} \ge \text{day of birth}$ | $\text{year of event} - \text{year of birth}$ |
| $\text{month of event} = \text{month of birth}$, $\text{day of event} \lt \text{day of birth}$ | $\text{year of event} - \text{year of birth} - 1$ |
| $\text{month of event} < \text{month of birth}$ | $\text{year of event} - \text{year of birth} - 1$ |



```{r}
dth_age_yr <- deaths
dth_age_yr$age_10 <- date_to_age_year(date = dth_age_yr$date_death,
                                      dob = dth_age_yr$date_birth)
dth_age_yr
```


```{r}
dth_age_month <- deaths
dth_age_month$age <- date_to_age_month(date = dth_age_month$date_death,
                                       dob = dth_age_month$date_birth)
dth_age_month
```
```{r}
dth_age_quarter <- deaths
dth_age_quarter$age <- date_to_age_quarter(date = dth_age_quarter$date_death,
                                           dob = dth_age_quarter$date_birth)
dth_age_quarter
```



## Cohorts

same as periods

```{r}
deaths_temp <- deaths
deaths_temp$cohort <- date_to_cohort_year(date = deaths_temp$date_birth)
deaths_temp
```

```{r, echo = FALSE, fig.height = 1.75}
demprep:::plot_date_to_cohort_year(date = deaths_temp$date_birth)
```



# Lexis triangles

- age and time use same length steps (subject to complications of months and years having different lengths, plus open age group)
- one month; one quarter; one year; $n$ years
- age and time not enough to identify cohort; also need Lexis triangle
- abstract: 
- event is in upper Lexis triangle

- An event occurring during period $t$ to a person in age group $a$ belongs to the lower Lexis triangle if the person attains age $a$ during, and to the upper Lexis triangle otherwise. (Covers youngest and oldest age groups.)


- If index of time equals index for age, then Lower; else Upper.

How to work out triangle:

1. Use date now and dob to calculate age group now
2. Use date now to identify date at start of period
3. Use date at start of period and dob to calculate whether entered age group during period
4. If so, lower; if not upper.



```{r}
dth_tri_yr <- deaths
dth_tri_yr$tri_jan <- date_to_triangle_year(date = dth_age_yr$date_death,
                                            dob = dth_age_yr$date_birth)
dth_tri_yr
```

```{r, echo = FALSE, fig.height = 6.5}
demprep:::plot_date_to_triangle_year(date = dth_tri_yr$date_death,
                                     dob = dth_tri_yr$date_birth)
```

Change from 1 Jan to 1 Jul

```{r}
dth_tri_yr$tri_jul <- date_to_triangle_year(date = dth_age_yr$date_death,
                                            dob = dth_age_yr$date_birth,
                                            month_start = "Jul")
dth_tri_yr
```

```{r}
dth_tri_qu <- deaths
dth_tri_qu$tri <- date_to_triangle_quarter(date = dth_tri_qu$date_death,
                                           dob = dth_tri_qu$date_birth)
dth_tri_qu
```

```{r}
dth_tri_month <- deaths
dth_tri_month$tri <- date_to_triangle_month(date = dth_tri_month$date_death,
                                            dob = dth_tri_month$date_birth)
dth_tri_month
```



# Cleaning existing labels


# Creating final labels

| Periods | Cohorts | Age groups | Lexis triangles |
|:--------|:--------|:-----------|:----------------|
| `format_period_year` | `format_cohort_year` | `format_age_year` | `format_triangle_year` |
| `format_period_multi` | `format_cohort_multi` | `format_age_multi` | `format_triangle_multi` |
| `format_period_custom` | `format_cohort_custom` | `format_age_custom` | \<none\> |
| \<none\> | \<none\> | `format_age_lifetab`| \<none\> |
| \<none\> | \<none\> | `format_age_births` | `format_triangle_births` |
| `format_period_quarter` | `format_cohort_quarter` | `format_age_quarter` | `format_triangle_quarter` |
| `format_period_month` | `format_cohort_month` | `format_age_month` | `format_triangle_month` |



# Tabulation